---
title: "Reproducible Research - Assignment 1"
author: "Corinne Helman"
date: "April 27, 2017"
output: 
#    html_document:
    md_document:
        variant: markdown_github
---
# loading knitr package
require(knitr)

# making sure that code and results are printed 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# setting proper directory
```{r load file}
setwd("C:/Users/helmac1/Documents/Personal/Coursera/Class 5 Programming Assignment 1")

# Read in the data set for the assignment
activity <- read.csv(file="C:/Users/helmac1/Documents/Personal/Coursera/Class 5 Programming Assignment 1/activity.csv", header=TRUE, sep=",")
summary(activity)
raw.activity <-read.csv(file="C:/Users/helmac1/Documents/Personal/Coursera/Class 5 Programming Assignment 1/activity.csv", header=TRUE, sep=",") 
```

# Transform the date column into a proper Date class
```{r clean date}
activity <- transform(activity, date = as.Date(date))
str(activity)
```
# Remove NA values
```{r}
activity <- activity[complete.cases(activity),]
```
# Calculate Total Number of Steps per Day
```{r}
totalStepsPerDay <- aggregate(steps ~ date, activity, sum)
```
# plot histogram of total number of steps per day
```{r plot steps}
hist(totalStepsPerDay$steps, breaks=20, xlab= "Total Number of Steps per day", ylab= "Frequency", main="Histogram of the total number of steps taken per day",col="blue", xlim=c(0,25000), ylim=c(0,12) )
``` 
# calculate and print mean and median
```{r mean and median}
steps.mean <- mean(totalStepsPerDay$steps)
steps.median <- quantile(totalStepsPerDay$steps, 0.5)

rbind("The Mean of the total number of steps per day", print(steps.mean))
rbind("The Median of the total number of steps per day", print(steps.median))
```
#plot daily activity patterns
```{r}
plot(activity$interval, activity$steps, type="l", col="blue", xlab="5 Minute Intervals", ylab="Average Number of Steps")
```
#Which 5 minute interval contains the max number of steps across all days
```{r}
averageStepsPerInterval <- aggregate(steps ~ interval, activity, mean)
rbind("The interval with the maximum number of steps", averageStepsPerInterval[which.max(averageStepsPerInterval$steps),1])
```
# Counting missing values
```{r}
countNA <- length(raw.activity$steps)-length(activity$steps)
rbind("The number of rows with missing data is",countNA)
```
# Substitute Missing Data with mean value of steps
```{r}
mean(averageStepsPerInterval$steps)
impute.activity <- raw.activity
impute.activity[is.na(impute.activity)]<-mean(averageStepsPerInterval$steps)
```
#Histogram of total number of daily steps
```{r}
imptotalStepsPerDay <- aggregate(steps ~ date, impute.activity, sum)
hist(imptotalStepsPerDay$steps, breaks=20, xlab= "Total Number of Steps per day", ylab= "Frequency", main="Histogram of the total number of steps taken per day",col="blue", xlim=c(0,25000), ylim=c(0,20) )
```
# Variance in activity between weekday and weekends
```{r}
impute.activity <- transform(impute.activity, date = as.Date(date))
str(impute.activity)
impute.activity$days <- weekdays(impute.activity$date)
impute.activity$days[impute.activity$days == "Monday"] <- "Weekday"
impute.activity$days[impute.activity$days == "Tuesday"] <- "Weekday"    
impute.activity$days[impute.activity$days == "Wednesday"] <- "Weekday"
impute.activity$days[impute.activity$days == "Thursday"] <- "Weekday"    
impute.activity$days[impute.activity$days == "Friday"] <- "Weekday"
impute.activity$days[impute.activity$days == "Saturday"] <- "Weekend"    
impute.activity$days[impute.activity$days == "Sunday"] <- "Weekend"

weekdaymat <-subset(impute.activity, impute.activity$days == "Weekday")
weekendmat <- subset(impute.activity, impute.activity$days == "Weekend")

weekdayStepsPerInterval <- aggregate(steps ~ interval, weekdaymat, mean)
weekendStepsPerInterval <- aggregate(steps ~ interval, weekendmat, mean)

plot(weekdayStepsPerInterval$interval, weekdayStepsPerInterval$steps, type="l", col="blue", xlab="5 Minute Intervals", ylab="Average Number of Steps during the Weekdays", main = "Weekdays")
plot(weekendStepsPerInterval$interval, weekendStepsPerInterval$steps, type="l", col="blue", xlab="5 Minute Intervals", ylab="Average Number of Steps during the Weekends", main = "Weekends")
```
